---
title: "Hard-to-sample species are more sensitive to land-use change: implications for global biodiversity metrics"
author: "Claudia Gutierrez-Arellano, Tim Newbold, Jenny Hodgson"
date: '`r format(Sys.time(), "%B, %Y")`'

output:
  html_document: 
    collapsed: false
    theme: flatly
    highlight: kate
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: false
  github_document:
    toc: yes
    toc_depth: 4
    html_preview: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, tidy = TRUE) 

```


This document compiles the code used to obtain the 'Records model' and results of the article "Hard-to-sample species are more sensitive to land-use change: implications for global biodiversity metrics".  

## 1. Data
We used the number of occurrence records in the international collaborative database [GBIF (Global Biodiversity Information Facility)](https://www.gbif.org/) and land cover type and intensity-related abundance data compiled in the [PREDICTS (Projecting Responses of Ecological Diversity in Changing Terrestrial Systems)](https://data.nhm.ac.uk/dataset/release-of-data-added-to-the-predicts-database-november-2022) database.

### 1.1 GBIF occureence count

We obtained the number of records (i.e. occurrence count) of all bird, plant and spider species recorded in GBIF. Below, the example code to obtain occurrence count of bird species.

```{r , eval=FALSE}
#Load necessary packages
library(tidyverse) #set of packages to facilitate data manipulation
library(rgbif) #access to data from GBIF (https://www.gbif.org/) via their API

#Bird species checklist from GBIF
gbif_birds<-readRDS('data/gbif_spp_birds.rds')

#Filter valid cases 
birds <- gbif_birds %>%
  #blank cells are transformed to 'NA'
  mutate_all(~ na_if(., ""))%>%
  #drop cases where 'genus' and 'species' fields are NA
  drop_na(genus, species)%>%
  #select unique cases for accepted taxon to remove possible synonyms
  distinct(acceptedTaxonKey, .keep_all = TRUE)%>%
  #filter 'accepted' taxa
  filter(taxonomicStatus=='ACCEPTED')%>%
  #filter records where the recognised taxon rank is 'species'
  filter(taxonRank=='SPECIES')%>%
  #filter extinct species
  filter(iucnRedListCategory!='EX'|iucnRedListCategory!='EW')

#Obtain occurrence number 
bird_checklist<- as.vector(birds$species)
bird_spp_records<- data.frame()
for (i in bird_checklist){
  #Get number of occurrence records 
  occ_count<- as.data.frame(occ_count(scientificName = i, year="1600,2023"))
  names(occ_count)<-'no_records'
  occ_count$species<-i
  occ_count$group<-paste0('Bird')
  bird_spp_records<-rbind(bird_spp_records, occ_count)
}

#Add taxonomic information
bird_occ_count<- merge(birds%>%select(species, taxonKey, class,order, family), 
                       bird_spp_records, by = "species", all.x = TRUE)


```

We repeated this for plant and spider species and compile the records for the three groups.

```{r , eval=FALSE}
#Join data frames of number of records per group 
all_groups_count<-rbind(bird_occ_count, plant_occ_count, spider_occ_count)%>%
  #Rename the 'species' column as Best_guess_binomial to relate to Predicts records
  rename(Best_guess_binomial= species)%>%
  #Filter species for which number of records is available and is not zero
  filter(!is.na(no_records))%>%
  filter(no_records>0)

```

### 1.2 PREDICTS databse

We obtained the land cover type and intensity-related abundance data compiled in the the latest version available of [PREDICTS (Contu et al., 2022; Hudson et al., 2016)](https://data.nhm.ac.uk/dataset/release-of-data-added-to-the-predicts-database-november-2022).

 PREDICTS contains around 4.3 millions of observations from studies assessing the relationship between land use type and intensity and several diversity metrics (e.g. abundance, species richness and biomass among others). Since our interest is focused on comparing change in local abundance per species we filtered the observations relevant to our aim.

#### 1.2.1 Filtered PREDICTS

We filtered: a) records where the _'Diversity metric'_ of the study was 'abundance', b) the specific epithet is known in the _'Best guess binomial'_ column (i.e. we excluded ‘sp.’ names), and c) we selected studies that assessed two or more of the (simpler) land use types. We simplified the land use categorisation of PREDICTS database by following the classification suggested by [Outhwaite et al 2022](https://doi.org/10.1038/s41586-022-04644-x)) 


```{r, eval=FALSE}
# Load necessary packages
library(tidyverse)  #set of packages to facilitate data manipulation

#Load latest PREDICTS database (2023)
predicts<- readRDS("data/predicts_updated_2023.rds")

#check if there are any blanks 
any(predicts == "")

#change any blanks to NAs
predicts <- predicts %>%
  mutate_all(~ na_if(., ""))

#Filter only 'Abundance' diversity metric type
incl_methods<-with(predicts,Diversity_metric_type=="Abundance" & Diversity_metric != "occurrence frequency")
table(incl_methods,exclude=NA)
predicts0<-predicts[incl_methods,]

#prepare suitable taxa
spotential<- !(is.na(predicts0$Family) | is.na(predicts0$Best_guess_binomial))

#ignore all 'Best_guess_binomial' with sp instead of species epithet
ss<-regexpr(" sp",as.character(predicts0$Best_guess_binomial),ignore.case=T)
ls<-nchar(as.character(predicts0$Best_guess_binomial))
summary(ls)#max(36)
hist(ls-ss)
unique(predicts0$Best_guess_binomial[(ls-ss)<5])

spotential<- spotential & (ls-ss)>4 

table(spotential,exclude=NULL)
table(is.na(spotential))#no NAs remaining

#simplify the database to sites only, but retaining some info about taxa included
psites<-predicts0%>%
  select(1:48)%>%
  distinct()

sapply(psites,function(x){length(unique(x))})# SSS and SSBS are same dimension as data frame, meaning everything else is invariant within them 

plistlengths<- predicts0 %>%
   select(SSS, 49, 59:64) %>%
   group_by(SSS) %>%
   summarise(across(everything(), ~length(unique(.))))


propidable<-tapply(spotential,INDEX=predicts0[,"SSS"],FUN=mean)

psites$propidable<-propidable[psites[,"SSS"]]

psites<-merge(psites,plistlengths,by="SSS",all=T)
with(psites,hist(propidable))
with(psites,hist(Genus))
with(psites,hist(Species))
with(psites,hist(Taxon_number))
with(psites,plot(Taxon_number+1,propidable,pch=4,log="x"))
#remove sites with no idable species
psites<-psites[psites$propidable>0,]

length(unique(psites$SS))
 

sort(-table(psites$Study_common_taxon))#quick look at taxa that have been retained

```
Below we explore representation of land uses, intensities, continents and biomes. 
```{r, eval=FALSE}
with(psites,table(Predominant_land_use,Use_intensity))

with(psites,table(Biome,exclude=NULL))

with(psites,table(UN_subregion,exclude=NULL))
with(psites,table(UN_region,exclude=NULL))


itab<-with(psites,table(Predominant_land_use,Use_intensity))

itab[,4]/rowSums(itab)  
#for secondary young, mature and undetermined use, >20% of sites' intensity cannot be decided

#create continent factor, intending to combine this with biome

#North america separated out so that temperate biomes in N and S america potentially have different properties

psites$Continent<-as.character(psites$UN_region)

psites$Continent[psites$UN_region=="Americas"]<-"America other"
psites$Continent[psites$UN_subregion=="North America"]<-"North America"

psites$Continent<-factor(psites$Continent)

table(psites$Continent)

#examine land-use and intensity differences among geographic areas

nsites<-aggregate(psites$Species,by=psites[,c("Predominant_land_use","Use_intensity","Continent", "Biome")],FUN=length)
nsites$Use<-with(nsites,factor(paste(Predominant_land_use,Use_intensity),levels=names(sort(-table(paste(Predominant_land_use,Use_intensity))))))
nsites$Place<-with(nsites,factor(paste(Continent, Biome),levels=names(sort(-table(paste(Continent, Biome))))))
habmod<-glm(x~Use*Biome +Use*Continent-1,data=nsites,family="poisson")
anova(habmod,test = "Chisq")
habsum<-summary(habmod)$coefficients
write.csv(habsum, "summary_data/landuse distribution_v2.csv")


##what secondary veg categories are available when primary veg is not?

primstudies<-unique(psites$SS[psites$Predominant_land_use=="Primary vegetation"])

length(primstudies)
length(unique(psites$SS))
#363/638 studies have a primary site for comparison, i.e. 275 do not

table(psites$Predominant_land_use[!(psites$SS %in% primstudies)])


matstudies<-unique(psites$SS[psites$Predominant_land_use %in% c("Intermediate secondary vegetation", "Mature secondary vegetation")])

length(matstudies)
sum(!(matstudies %in% primstudies))

638-363-101
#174 studies would have no comparison to Primary or inter/mature secondary

secstudies<-unique(psites$SS[psites$Predominant_land_use %in% c("Intermediate secondary vegetation", "Mature secondary vegetation","Young secondary vegetation","Secondary vegetation (indeterminate age)" )])

length(secstudies)
sum(!(secstudies %in% primstudies))

638-363-178
#97 studies would have no comparison to Primary or any secondary

itab
```
We now add the column _'Outhwaite'_, where we reclassify the observations' land use type (_'Predominant_land_use'_) and intensity (_'Use_intensity'_) following the classification suggested by [Outhwaite et al 2022](https://doi.org/10.1038/s41586-022-04644-x)).

```{r, eval=FALSE}
psites$Outhwaite<- as.character(psites$Predominant_land_use)

#Exclude urban and cannot decide
psites$Outhwaite[as.character(psites$Predominant_land_use) %in% 
                       c("Urban","Cannot decide")]<-NA
table(psites$Outhwaite)

#Aggregate all types of secondary vegetation
psites$Outhwaite[as.character(psites$Predominant_land_use) %in% 
                       c("Intermediate secondary vegetation", "Mature secondary vegetation","Young secondary vegetation","Secondary vegetation (indeterminate age)" )  ]<-"Secondary vegetation"

#Low-Intensity agriculture
psites$Outhwaite[as.character(psites$Predominant_land_use) %in% c("Plantation forest") &       psites$Use_intensity %in% c("Minimal use")]<- 'Low-intensity agriculture'

psites$Outhwaite[as.character(psites$Predominant_land_use) %in% c("Pasture") &       psites$Use_intensity %in% c("Minimal use", "Light use")]<- 'Low-intensity agriculture'

psites$Outhwaite[as.character(psites$Predominant_land_use) %in% c("Cropland") &       psites$Use_intensity %in% c("Minimal use", "Light use")]<- 'Low-intensity agriculture'

#Low-Intensity agriculture
psites$Outhwaite[as.character(psites$Predominant_land_use) %in% c("Plantation forest") &       psites$Use_intensity %in% c("Light use", "Intense use")]<- 'High-intensity agriculture'

psites$Outhwaite[as.character(psites$Predominant_land_use) %in% c("Pasture") &       psites$Use_intensity %in% c("Intense use")]<- 'High-intensity agriculture'

psites$Outhwaite[as.character(psites$Predominant_land_use) %in% c("Cropland") &       psites$Use_intensity %in% c("Intense use")]<- 'High-intensity agriculture'

#Exclude cases where uses have undecided intensity
psites$Outhwaite[as.character(psites$Predominant_land_use) %in% c("Cropland","Plantation forest","Pasture") & psites$Use_intensity %in% c("Cannot decide")]<-NA

#now remove sites with NA Use and studies that no longer have 2 different comparison land covers

names(psites)

psites0<-psites[!is.na(psites$Outhwaite),]

nuses<-tapply(psites0$Outhwaite, psites0$SS,function(x){length(unique(x))})

table(nuses)
# nuses Outhwaite classification
# 1   2   3   4 
# 211 271 124  18
#we lose 211 studies which don't have a comparison now 

sum(table(nuses))
#Outwhite
#624, 14 studies were dropped because contained wholly ineligible use categories

onelevel<-names(nuses[!is.na(nuses) & nuses==1])

with(psites0[psites0$SS %in% onelevel,],table(Predominant_land_use,Use_intensity))

with(psites0[psites0$SS %in% onelevel,],table(Biome,Continent))

with(psites0[psites0$SS %in% onelevel,],table(Predominant_land_use,Use_intensity))/with(psites,table(Predominant_land_use,Use_intensity))

#mainly primary and secondary vegetation sites have been lost, as they must have been compared within themselves
#about a quarter of pasture sites have been lost, and 20% of minimal use cropland

with(psites0[psites0$SS %in% onelevel,],table(Biome,Continent))/with(psites,table(Biome,Continent))
#many biomes and continents have a heavy loss of sites

psites0<-psites0[!(psites0$SS %in% onelevel),]

#check taxonomic groups in remaining sites

tt<-table(psites0$Study_common_taxon)

sort(tt[tt>0])#birds, plants and spiders still look viable 

####simplify biomes to broad classes of human impact###

table(psites0$Biome)

levels(psites0$Biome)

psites0$Impact_zone<-"Later impact"

#will include "Tropical & Subtropical Coniferous Forests" "Mangroves"  "Tropical & Subtropical Grasslands, Savannas & Shrublands"  "Tropical & Subtropical Dry Broadleaf Forests" 

psites0$Impact_zone[psites0$Biome %in% c("Temperate Broadleaf & Mixed Forests","Mediterranean Forests, Woodlands & Scrub" ,"Temperate Grasslands, Savannas & Shrublands")]<-"Old impact"

psites0$Impact_zone[psites0$Biome %in% c("Tundra" , "Boreal Forests/Taiga","Temperate Conifer Forests" ,
                                         "Deserts & Xeric Shrublands","Montane Grasslands & Shrublands"  )]<-"Low impact"

psites0$Impact_zone[psites0$Biome %in% c("Tropical & Subtropical Moist Broadleaf Forests")]<-"Moist Broadleaf"


psites0$Impact_zone<-factor(psites0$Impact_zone,levels=c("Low impact","Moist Broadleaf","Later impact","Old impact"))

table(psites0$Impact_zone)
# Low impact Moist Broadleaf    Later impact      Old impact 
# 1778            5648            2711            9502 


# match back to species data
names(psites0)
names(predicts0)
predicts0$Idable<-spotential

#we won't retain the columns 49:70 which counted the number of unique values at different taxon levels, because this won't stay the same

psites0$sitematch<-as.character(psites0$SSS)
predicts0$sitematch<-as.character(predicts0$SSS)

predicts1<-left_join(predicts0, psites0%>%select(49, 57:60), by="sitematch")%>%
  drop_na(Outhwaite)

#now remove species that don't occur in the selected sites

stem(predicts1$Measurement)

summary(predicts1$Idable)

#1.8 M idable

spsmax<-with(predicts1,tapply(Measurement,paste(SS,Taxon_number),max))
#38717 sp-study combinations

table(spsmax==0,exclude=NULL)
#1751 to exclude

summary(spsmax)

#investigate distributions of abundances 
predicts1$Tax_study_maxabund<-as.vector(spsmax[paste(predicts1$SS,predicts1$Taxon_number)])

predicts1[predicts1$Measurement>140000,]
#few extreme high abundances come from a study of nematodes per m^2 in New Zealand; none are idable

predicts1<-predicts1[predicts1$Idable & predicts1$Tax_study_maxabund>0,]

with(predicts1,min(Measurement[Measurement>0]))
#[1] 0.000625

hist(log10(predicts1$Measurement+0.000625/2))

studyminpos<-with(predicts1,tapply(Measurement,SS,function(x){min(x[x>0])}))

studyminpos<-studyminpos[names(studyminpos) %in% as.character(predicts1$SS)]#NAs were caused by SS factor levels no longer there

summary(studyminpos)#no NAs left

predicts1$Study_minpos<-as.vector(studyminpos[as.character(predicts1$SS)])

saveRDS(predicts1, 'data/predicts_updated_filtrd_Outhwaite.rds')

```


#### 1.2.2 Total abundance and effort

For each PREDICTS study, we calculated the total abundance per species at each land use type (i.e. summing across any sites with the same land use _sensu_ [Outhwaite et al 2022](https://doi.org/10.1038/s41586-022-04644-x)). To deal with zeros we added the minimum abundance per study to the total abundance. Also, we calculated the total effort per land cover type per study

```{r , eval=FALSE}

predicts1<- readRDS("data/predicts_updated_filtrd_Outhwaite.rds")

glimpse(predicts1)

predicts2 <- predicts1%>%
  #filter groups of interest
  filter(Class=='Aves'|Phylum =='Tracheophyta'|Order=='Araneae')%>% 
  #create a column to analyse groups separately
  mutate(group = ifelse(Class == 'Aves', 'Bird', NA))%>% 
  mutate(group = ifelse(Phylum == 'Tracheophyta', 'Plant', group))%>%
  mutate(group = ifelse(Order == 'Araneae', 'Spider', group))%>%
  #select relevant columns
  select(SS, group, Best_guess_binomial, Outhwaite, 
         Measurement, Rescaled_sampling_effort)%>%
  #sum of abundance and effort by study, species and land use, and create column with the number of sites added up
  group_by(SS, Best_guess_binomial, Outhwaite)%>%
  mutate(Measurement_total= sum(Measurement),
         effort_total= sum(Rescaled_sampling_effort),
         no_sites = n())%>% 
  #add up the minimum abundance per study to the total abundance
  group_by(SS)%>%
  #create column with the minimum
  mutate(Measurement_nozeros= Measurement_total + min(Measurement[Measurement> 0]), 
         min_Measurement= min(Measurement[Measurement> 0]))%>% 
  ungroup()

#Order land use categories for statistical analyses
predicts2 <- predicts2 %>%
  mutate(Outhwaite = factor(Outhwaite,
                            levels = c('Primary vegetation', 'Secondary vegetation', 'Low-intensity agriculture', 'High-intensity agriculture'),
                            ordered = TRUE))

```
We then filtered unique observations of species' total abundance per study and kept those where the species were recorded in more than one land use type

```{r , eval=FALSE}

#Filter unique observations of species aggregation per study 
predicts2<- predicts2%>%
  distinct(SS,group, Best_guess_binomial, Outhwaite, Measurement_nozeros,effort_total, no_sites, min_Measurement, studyspecies)
#Filter observations with more than one land use type
predicts_unique<-predicts2%>%
  group_by(Best_guess_binomial) %>%
  filter(n_distinct(Outhwaite) > 1)

```
#### 1.2.2 Adding and transforming number of records
We joined the data frame of number of records to the total abundance data frame 

```{r , eval=FALSE}
#Relate data frames of number of records and total abundance by species name 
predicts_occ_count<-predicts_unique %>%
  left_join(all_groups_count%>%select(Best_guess_binomial, no_records,class, order, family), by = "Best_guess_binomial")

```
We then log10-transformed the number of records and re-scaled them by subtracting the mean of this sample (_ctrlogrec_).

```{r , eval=FALSE}
#Calculate the mean
rec_log10mean<-mean(log10(predicts_occ_count$no_records),na.rm = TRUE)

#Re-scale log10-transformed number of records and add as a new variable
predicts_occ_count$ctrlogrec<- log10(predicts_occ_count$no_records) - rec_log10mean

```

## 2. Analyses

### 2.1 Records model

We produced a linear mixed-effect model using the total abundance (_Measurment_nonzeros_) as  the response variable, land cover type (_Outhwaite_), taxonomic group (_group_) and the interaction between group and the transformed number of species records (_ctrlogrec_) as fixed effects.  Study, and species within study as random effects, and offset the log10-transformed total effort per land cover type per study. 

```{r , eval=FALSE}

#Load necessary packages
library(lme4)  #linear model with random effects 
library(lmerTest) #provides p-values in summary tables for lme4 models
records_model<-lmer(log10(Measurement_nozeros)~ Outhwaite*(ctrlogrec:group)+ group +(1|SS/Best_guess_binomial) + offset(log10(effort_total)), predicts_occ_count, na.action='na.omit')

summary(records_model)
```

#### 2.1.1 Models comparison
We compared the goodness-of-fit of alternative structure of fixed effects and Records model had the lowest AIC. 
```{r, eval=FALSE}

#model without number of records
rm_b<-lmer(log10(Measurement_nozeros)~ Outhwaite*group+ group +(1|SS/Best_guess_binomial) + offset(log10(effort_total)), predicts_occ_count, na.action='na.omit')
summary(rm_b)

#model without interaction
rm_c<-lmer(log10(Measurement_nozeros)~ Outhwaite+(ctrlogrec:group)+ group +(1|SS/Best_guess_binomial) + offset(log10(effort_total)), predicts_occ_count, na.action='na.omit')
summary(rm_c)

anova(records_model,rm_b,rm_c)


```


#### 2.1.2 Land use chage effect on abundance

```{r, eval=FALSE}

#Polynomial contrast matrix for four land use types
shape<-contr.poly(4)

# Contrast coefficients * Records model estimates
-0.2242*shape[,1]-0.03557*shape[,2]-0.02542*shape[,3]
#  0.13829702  0.05086539 -0.01529539 -0.17386702

# calculate the difference between levels
diff(-0.2242*shape[,1]-0.03557*shape[,2]-0.02542*shape[,3])
# -0.08743163 -0.06616078 -0.15857163

#plot 
x<-1:4
y<- -0.2242*shape1[,1]-0.03557*shape1[,2]-0.02542*shape1[,3]

plot(x, y, pch = 19, xlab = 'Land use', ylab = 'Coefficient* Estimate', xaxt = "n", main = 'Land use effect on abundance', cex.main = 0.9 )
axis(1, at= 1:4,labels = c("Primary \n vegetation", "Secondary \nvegetation", "Low-intensity \nagriculture",  "High-intensity\n agriculture"), cex.main = 1, cex.axis = 0.8)
lines(x,y)

```


### 2.2 Model predictions

We used the Records model (_records_model_) to predict per species abundance at each land use type.

```{r , eval=FALSE}
#To predict species abundance with the Records model we must define all possible levels of land use types, groups and number of records 

#Select relevant columns to define the observed records sample
sample_records <- predicts_occ_count %>%
  distinct(group, Best_guess_binomial, ctrlogrec)


# Obtain for all possible levels of number of records, groups and land use types
predicts_dummy<- sample_records %>%
  group_by(group, ctrlogrec) %>%
  mutate(nspp=n()) %>%
  distinct(group, ctrlogrec,nspp) %>%
  mutate(effort_total = 1)

Outhwaite <- predicts_occ_count%>%
  ungroup()%>%
  distinct(expand.grid(predicts_occ_count$Outhwaite))%>%
  rename(Outhwaite = Var1)

predicts_dummy_s <- crossing(predicts_dummy, Outhwaite) %>%
  ungroup() 

# define LU and group as factor variables
predicts_dummy_s$Outhwaite<- factor(predicts_dummy_s$Outhwaite, levels = c("Primary vegetation", "Secondary vegetation",
                   "Low-intensity agriculture", "High-intensity agriculture"), ordered = TRUE)
predicts_dummy_s$group<-factor(predicts_dummy_s$group)

#Model predictions
all_predicted <- predict(records_model, predicts_dummy_s, re.form = NA)


# Organise predictions one column per land use
pred_s_RM <- predicts_dummy_s %>%
  #add predicted values
  mutate(predicted = all_predicted) %>%
  pivot_wider(id_cols = c(group, ctrlogrec,nspp), names_from = c(Outhwaite),
              values_from = c(predicted), names_sort = TRUE) %>%
  ungroup()

```

#### 2.2.1 Proportional differences
We calculated the proportional difference as the geometric mean difference between one land-use and another across species.

```{r, eval=FALSE}

# Calculate all differences from primary vegetation

Onames<-levels(predicts_dummy_s$Outhwaite)

pred_s_RM[,Onames]<-sapply(pred_s_RM[,Onames],function(x){x-pred_s_RM[,Onames[1]]})

# Calculate weighted mean
xx<- base::lapply(pred_s_RM[,Onames], function(a){weighted.mean(unlist(a),w=unlist(pred_s_RM[,"nspp"]))})

wtmeans_s<-as.data.frame(pred_s_RM)[0,Onames]

wtmeans_s<-rbind(wtmeans_s,xx)

#proportional differences
prop_diff<-as.data.frame(10^apply(wtmeans_s[,2:4],2,mean)-1)
prop_diff$Outhwite<-rownames(prop_diff)
names(prop_diff) <- c("prop_diff", "Outhwaite")
prop_diff$obs<- 'Predicts'

```             
#### 2.2.2 Confidence intervals
We placed confidence intervals on the predicted, transformed summary of data , i.e. the geometric mean difference between one land-use and another across species (in  _prop_diff_). We accounted for uncertainty in all the model fixed effects and their intercorrelations.

Our approach generates large populations of complete sets of plausible model parameters, as it is more efficient than working out the right way to combine the CIs on each individual parameter. The 'confidence' of this confidence interval should be understood as the confidence in the average effect, not the confidence of what we would find sampling any one location or species

```{r, eval=FALSE}
library(lme4)  #linear model with random effects 
library(lmerTest)  #provides p-values in summary tables for lme4 models
library(tidyverse) #data managment 
library(Matrix)
library(gdata)#for upperTriangle of matrices
library(MASS)#for principal components

# Obtain model parameter covariance matrix ----
covmat<-vcov(records_model)

# Convert it to a correlation matrix
cormat<-cov2cor(covmat)

stabs<-summary(records_model)$coefficients # the significance table

# Verify that the variances calculated from covmat match the squared standard errors provided in the model summary stabs
plot(diag(covmat),stabs[,2]^2,log="xy")
text(diag(covmat),stabs[,2]^2,rownames(stabs),pos=4)

# Range of correlations observed
hist(log10(abs(upperTriangle(cormat))*100),breaks=50)# a small tail below 10^-2

image(log10(abs(cormat)*100),breaks=c(-5,-2,-1,0,1,2))


# Generate matrix of random normal variables ----

uncor<-matrix(rnorm(100000),nrow=5000) # 20 variables

# standardise variables
uncor<-as.matrix(scale(uncor))

# compute and visualise correlations
hist(upperTriangle(cor(uncor))*100,breaks=30)# correlations exist, most in the range -2...2%

# compute principal component analysis
pcuc<-princomp(uncor)

# subset and scale principal components
uncor2<-pcuc$scores[,1:18]
uncor2<-as.matrix(scale(uncor2))

# check effect of scaling 
apply(uncor2,2,mean)# scaling worked
apply(uncor2,2,var)

hist(upperTriangle(cor(uncor2))*100,breaks=30)# minute remaining correlations

# Generate correlated variables ----

hist(log10(abs(upperTriangle(covmat,diag=T))))

# Range over 7 ord. magnitude. In case improves precision, scale up
covbig<-covmat*10^4

# Cholesky decomposition matrix
cholbig<-chol(covbig)

# Compute covariance and correlation of the standardised principal components transformed with the decomposition matrix
cored1<- t(as.matrix(t(cholbig)) %*% t(uncor2))
covout1<-cov(cored1)
corout1<-cor(cored1)

# Comparison between original and transformed covariance and correlation matrices
plot(upperTriangle(covmat),upperTriangle(covout1),pch=4) 

plot(abs(upperTriangle(cormat)),abs(upperTriangle(corout1)),log="xy",pch=4)
abline(0,1)


# Generate coefficient population dataframe ----

# Give variables the correct mean and variance 
coefpop<-as.data.frame(cored1)
names(coefpop)<-rownames(stabs)

coefpop<-scale(coefpop,center=T,scale=apply(coefpop, 2, sd)/stabs[,2]) # center at zero to divide
coefpop<-scale(coefpop,center=-stabs[,1],scale=F)# then centre
cbind(apply(coefpop,2,mean),# scaling worked
apply(coefpop,2,sd))

```

Now we obtain the confidence intervals by generating new predictions based on the new sets of plausible model parameters (i.e. _coefpop_)

```{r, eval=FALSE}

# dummy response variable needed for model matrix added to dummy data created in section 2.2
predicts_dummy_s$Measurement_nozeros<- 10^all_predicted

# Generate model matrix for dummy data

mm<-model.matrix(log10(Measurement_nozeros)~ Outhwaite*(ctrlogrec:group)+ group + offset(log10(effort_total)),data=predicts_dummy_s)
dim(mm)

for(i in 1:5000){
  
  coefnow<-coefpop[i,]
  
  predicted <- mm %*% coefnow 
  
  pred_s_RM <- predicts_dummy_s %>%
    mutate(predicted = as.vector(predicted)) %>%
    pivot_wider(id_cols = c(group, ctrlogrec,nspp), names_from = c(Outhwaite),
                values_from = c(predicted), names_sort = TRUE) %>%
    ungroup()
  
  #calculate all differences from primary 
  pred_s_RM[,Onames]<-sapply(pred_s_RM[,Onames],function(x){x-pred_s_RM[,Onames[1]]})
  
  #weighted mean
  xx<-  as.data.frame(base::lapply(pred_s_RM[,Onames],function(a){weighted.mean(unlist(a),w=unlist(pred_s_RM[,"nspp"]))}))
  
  #add to previous
  wtmeans_s<-rbind(wtmeans_s,xx)
  
}

prop_diff_ci<-data.frame(matrix(ncol = 0, nrow = 3))
prop_diff_ci$Outhwaite<-names(wtmeans_s[,2:4])
prop_diff_ci$lowerc<-10^apply(wtmeans_s[,2:4],2,quantile,probs=0.025)-1
prop_diff_ci$upperc<-10^apply(wtmeans_s[,2:4],2,quantile,probs=0.975)-1

prop_diff<- merge(prop_diff, prop_diff_ci, by = 'Outhwaite')

```


### 2.3 Model extrapolation
Now we use the Records model to extrapolate predictions to bird, plant and spider species not included in the PREDICTS database for which number of records was available (using the same calculations as above, just with more species). 

See section [1.1 GBIF occurrence]((### 1.1 GBIF occurrence)) to obtain the _all_groups_count_ dataframe and section [1.2.2 Adding and transforming number of records]((#### 1.2.2 Adding and transforming number of records)) to obtain the mean value ( _rec_log10mean_ ) used to center the log10 transformed number of records. We then estimated the relative abundance change.

```{r, eval=FALSE}

# First we transform the number of records of all species available in GBIF

#In section 1.2.2
#rec_log10mean<-mean(log10(predicts_occ_count$no_records),na.rm = TRUE)

all_groups_count$ctrlogrec<- log10(all_groups_count$no_records)-rec_log10mean

# subset of no. of records available
available_records <- all_groups_count %>%
  mutate(group = ifelse(group == "Aves", "Bird", group)) %>%
  mutate(group = ifelse(group == "Tracheophyta", "Plant", group)) %>%
  mutate(group = ifelse(group == "Araneae", "Spider", group)) %>%
  distinct(group, Best_guess_binomial, ctrlogrec)

# data frame with ordered level factors of land use
Outhwaite <- predicts_occ_count %>%
  ungroup() %>%
  distinct(expand.grid(predicts_occ_count$Outhwaite)) %>%
  rename(Outhwaite = Var1)

# predict values for all possible levels of LU groups and number of records, retaining number of spp
predicts_dummy <- available_records %>%
  group_by(group, ctrlogrec) %>%
  mutate(nspp=n()) %>%
  distinct(group, ctrlogrec,nspp) %>%
  mutate(effort_total = 1)


predicts_dummy_extrap <- crossing(predicts_dummy, Outhwaite) %>%
  ungroup() 

# define LU and group as factor variables
predicts_dummy_extrap$Outhwaite<- factor(predicts_dummy_extrap$Outhwaite, levels = c("Primary vegetation", "Secondary vegetation", "Low-intensity agriculture", "High-intensity agriculture"), ordered = TRUE)

predicts_dummy_extrap$group<-factor(predicts_dummy_extrap$group)

# Model extrapolation predictions
all_predicted <- predict(records_model, predicts_dummy_extrap, re.form = NA)


# Organise predictions one column per land use
pred_extrap_RM <- predicts_dummy_extrap %>%
  mutate(predicted = all_predicted) %>%
  pivot_wider(id_cols = c(group, ctrlogrec,nspp), names_from = c(Outhwaite),
              values_from = c(predicted), names_sort = TRUE) %>%
  ungroup()

Onames<-levels(predicts_dummy_extrap$Outhwaite)

pred_extrap_RM[,Onames]<-sapply(pred_extrap_RM[,Onames],function(x){x-pred_extrap_RM[,Onames[1]]})

# Calculate weighted mean
xx<- base::lapply(pred_extrap_RM[,Onames], function(a){weighted.mean(unlist(a),w=unlist(pred_extrap_RM[,"nspp"]))})

wtmeans_x<-as.data.frame(pred_extrap_RM)[0,Onames]

wtmeans_x<-rbind(wtmeans_x,xx)

# Proportional differences
prop_diff_extrap<-as.data.frame(10^apply(wtmeans_x[,2:4],2,mean)-1)
prop_diff_extrap$Outhwite<-rownames(prop_diff)
names(prop_diff_extrap) <- c("prop_diff", "Outhwaite")
prop_diff_extrap$obs<- 'Extrapolated'


#CI for extrapolation estimates----

# dummy response variable needed for model matrix added to dummy data obtained above
predicts_dummy_extrap$Measurement_nozeros<- 10^all_predicted

# Generate model matrix for extrapolated dummy data

mm<-model.matrix(log10(Measurement_nozeros)~ Outhwaite*(ctrlogrec:group)+ group + offset(log10(effort_total)),data=predicts_dummy_extrap)
dim(mm)


for(i in 1:5000){
  
  coefnow<-coefpop[i,]
  
  predicted <- mm %*% coefnow 
  
  pred_extrap_RM <- predicts_dummy_extrap %>%
    mutate(predicted = as.vector(predicted)) %>%
    pivot_wider(id_cols = c(group, ctrlogrec,nspp), names_from = c(Outhwaite),
                values_from = c(predicted), names_sort = TRUE) %>%
    ungroup()
  
  #calculate all differences from primary 
  pred_extrap_RM[,Onames]<-sapply(pred_extrap_RM[,Onames],function(x){x-pred_extrap_RM[,Onames[1]]})
  
  #weighted mean
  xx<-  as.data.frame(base::lapply(pred_extrap_RM[,Onames],function(a){weighted.mean(unlist(a),w=unlist(pred_extrap_RM[,"nspp"]))}))
  
  #add to previous
  wtmeans_x<-rbind(wtmeans_x,xx)
  
}


prop_diff_extra_ci<-data.frame(matrix(ncol = 0, nrow = 3))
prop_diff_extra_ci$Outhwaite<-names(wtmeans_x[,2:4])
prop_diff_extra_ci$lowerc<-10^apply(wtmeans_x[,2:4],2,quantile,probs=0.025)-1
prop_diff_extra_ci$upperc<-10^apply(wtmeans_x[,2:4],2,quantile,probs=0.975)-1

prop_diff_extrap<- merge(prop_diff_extrap, prop_diff_extra_ci, by = 'Outhwaite')


```

## 3. Extrapolation robustness
We assessed the robustness of the 'Records model' to extrapolate predictions species not included in PREDICTS (i.e., species whose only information available is number of records obtained from GBIF).

First, we calculated the probability of a species being absent in at least two land uses as the proportion of zero abundance cases in the _predicts_occ_count_ database squared (0.322= 0.10). We used this probability as a threshold to divide the database into testing (N=2,017) and training datasets (N= 16,681).

We assessed the general fitness of the testing dataset (i.e. cases with less than 10% mean abundance by species per studies) to a) the Records model, (obtained with all observations) and b) the Training model, obtained with the remaining 90% of observations (i.e. cases with more than 10% mean abundance by species per studies).

### 3.1 Fitness to Records model

```{r, eval=FALSE}
#Calculate the mean abundance values per species by study
predicts_occ_mean_ab<- predicts_occ_count%>%
  mutate(trans_ab= log10(Measurement_nozeros))%>%
  group_by(SS, Best_guess_binomial)%>%
  summarize(mean_ab= mean(trans_ab))

#Studies and species in with the lowest 10% mean abundance
test_ss_spp<-predicts_occ_mean_ab%>%
  filter(ntile(mean_ab, 10) == 1)

#Split predicts_occ_count into testing and training based on the 10% lowest mean abundance

#Testing dataset
predicts_occ_test<- predicts_occ_count%>%
  inner_join(test_ss_spp, by = c("SS" = "SS", "Best_guess_binomial" = "Best_guess_binomial"))

length(predicts_occ_test$SS)/length(predicts_occ_count$SS)
#~11% of records included in testing data

#Training dataset
predicts_occ_train <- predicts_occ_count %>%
  ungroup()%>%
  anti_join(predicts_occ_test)

#Predict testing abundance
#subset of no. of records available
sample_records<-predicts_occ_test%>%
  distinct(group,Best_guess_binomial, ctrlogrec)

#dummy dataset with all possible combinations of land use, taxonomic group and given no. of records in the Testing dataset
predicts_dummy_rm_test<-predicts_occ_test%>%
  ungroup()%>%
  select(Outhwaite, group, ctrlogrec)%>%
  expand (Outhwaite, group, ctrlogrec)%>%
  mutate(effort_total = 1)

#predict abundance values with Records Model
all_predicted<- predict(records_model, predicts_dummy_rm_test, re.form=NA)

RM_test_predicted<- predicts_dummy_rm_test%>%
  mutate(predicted = all_predicted)

#Retain real combinations of group and no. of records
predicted_RM_test<-RM_test_predicted%>%
  inner_join(sample_records, RM_test_predicted, by = NULL)%>%
  distinct()


#Create a dataframe of observed vs predicted values of the testing data
test_predicts_obs_pred<- inner_join(predicts_occ_test, predicted_RM_test, by =c("ctrlogrec","Best_guess_binomial", 'Outhwaite'))

r_squared_a<-(cor(log10(test_predicts_obs_pred$Measurement_nozeros), test_predicts_obs_pred$predicted))^2
# 0.001878686

plot(test_predicts_obs_pred$predicted, log10(test_predicts_obs_pred$Measurement_nozeros),  main = 'Records model', xlab = "predicted", ylab = "observed" )
abline(a = 0, b = 1, col = "red", lty = 2)



```


### 3.2 Fitness to Training model (above 10% mean abundance)

 
```{r, eval=FALSE}

# Training model
RM_train<- lmer(log10(Measurement_nozeros)~ Outhwaite*(ctrlogrec:group)+ group +(1|SS/Best_guess_binomial) + offset(log10(effort_total)), predicts_occ_train, na.action='na.omit')

summary(RM_train)

# Predict abundance values with Training Model (dummy dataset obtained above with all possible combinations of land use, taxonomic group and given no. of records in the Testing dataset)
all_predicted<- predict(RM_train, predicts_dummy_rm_test, re.form=NA)

RM_test_predicted<- predicts_dummy_rm_test%>%
  mutate(predicted = all_predicted)

#Retain real combinations of group and no. of records
predicted_RM_test<-RM_test_predicted%>%
  inner_join(sample_records, RM_test_predicted, by = NULL)%>%
  distinct()

#Create a dataframe of observed vs predicted values of the testing data
test_predicts_obs_pred<- inner_join(predicts_occ_test, predicted_RM_test, by =c("ctrlogrec","Best_guess_binomial", 'Outhwaite'))

r_squared_b<-(cor(log10(test_predicts_obs_pred$Measurement_nozeros), test_predicts_obs_pred$predicted))^2
#0.0017

plot(test_predicts_obs_pred$predicted, log10(test_predicts_obs_pred$Measurement_nozeros),  main = 'Training model: lower 10% mean abundance', xlab = "predicted", ylab = "observed" )
abline(a = 0, b = 1, col = "red", lty = 2)

```



## 4. BII comparison
We compared our estimations of change with those obtained using the BII model. We obtained the BII estimates by following the [methods described by De Palma et al. (2024)](https://adrianadepalma.github.io/BII_tutorial), using the land cover classification proposed by [Outhwaite et al. 2022](https://doi.org/10.1038/s41586-022-04644-x) and the species _‘Best guess binomial’_ to calculate compositional similarity

We obtained the BII index for a) the same subset of PREDICTS studies we used to fit our model and b) all studies suitable for the index calculation (i.e. those with at least some Primary vegetation data and more than one species). 

